---
title: "Project 2"
author: "Jim og Christian"
date: "03 02 2022"
output: 
  bookdown::pdf_document2:
    
    number_sections: false
  # pdf_document:
  #   toc: no
  #   toc_depth: '2'
subtitle: Spatial Statistics
urlcolor: blue
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,
                      strip.white=TRUE,
                      prompt=FALSE,
                      cache=TRUE, 
                      size="scriptsize",
                      fig.width=5, 
                      fig.height=5, 
                      fig.align = "center")
```

```{r packages, options}
library(geoR)    
library(akima)   
library(fields)   
library(MASS)     
library(viridis)  
library(tidyverse)
library(spatstat)
library(scatterplot3d)
library(plotly)
```

```{r figplotly, include=F}
s3dPlot = function(grid, fn="surfacePlot.png",z=1.8, x=1, y=1){
  fig = plot_ly(x=grid$x, y=grid$y,z = grid$z) %>% 
    add_surface(
      colors = terrain.colors(5),
      contours = list(
        z = list(
          show=TRUE,
          usecolormap=F,
          highlightcolor="#ff1111",
          project=list(z=TRUE)
          ),
        start = 0,
        end=1000,
        size = 0
      )
    ) 
  m <- list(
    l = 1,
    r = 1,
    b = 1,
    t = 1,
    pad = 100
  )
  fig <- fig %>% layout(
    margin = m,
    scene = list(
      xaxis = list(title = "y"), 
      yaxis = list(title = "x"),
      zaxis = list(title = "Elevation"),
      camera=list(
        eye = list(x=x, y=y, z=z)
        )
      )
    )%>% 
    config(
      toImageButtonOptions = list(
        format = "png",
        filename = "myplot",
        width = 800,
        height = 700
      )
    )
  save_image(fig, file=fn)
  return(fig)
}
```

# 2 GRF - Real Data
## a) Visualization

Our dataset contain $n = 52$observations of terrain elevation located in the domain $\mathcal{D} = [0, 315]^2\subset\mathbb{R^2}$.

Visualization \textcolor{blue}{Vet ikke om d trengs noe mer intro til visualization?} 
```{r P2visualization, options}
figPath = "./Fysmat/8 Semester V2022/RomStat/romstat/project1/Problem2Figures/"

dt <- read.table(file="~/Fysmat/8 Semester V2022/RomStat/Project1/Prj1Code_romstat/topo.dat")

setwd("C:/Users/chris/OneDrive/Documents/Fysmat/8 Semester V2022/RomStat/romstat/project1/Problem2Figures/")

# Setup
gd = as.geodata(dt)
grid <- interp(dt$x, dt$y, dt$z)
griddf <- subset(data.frame(
  x = rep(grid$x, nrow(grid$z)),
  y = rep(grid$y, each = ncol(grid$z)),
  z = as.numeric(grid$z)), !is.na(z))
# pdf("2a_contour.pdf")
image.plot(grid, col=terrain.colors(200), asp=1,
  xlab="x", ylab="y"
)
contour(grid, add=T)
points(dt, pch=4, col="red", )
# dev.off()

# surface2a = s3dPlot(grid, fn="2a_surface.pdf")

# Linear regression to see linear data trend
# pdf("2a_linearPlane.pdf")
plane = lm(dt$z ~ dt$x + dt$y)
planeplot = scatterplot3d(dt$x,dt$y,dt$z, angle=-20,
  xlab = "x", ylab = "y", zlab = "z")
planeplot$plane3d(plane)
# dev.off()

pdf("2a_obsDensity.pdf")
heat <- as.ppp(cbind(dt$x,dt$y), c(0, 315, 0, 315))
plot(density(heat, adjust=0.5), col=viridis(200, direction=-1), main="")
dev.off()
# df = as.data.frame(dt)
```

```{r 2a_savefigsDelete, include=F, eval=F}
# pdf(2a_visualization)
# par(mfrow=(c(1,2)))
image.plot(grid, col=terrain.colors(200), asp=1,
  xlab="x", ylab="y"
)
contour(grid, add=T)
points(dt, pch=4, col="red", )
# dev.off()

plane = lm(dt$z ~ dt$x + dt$y)
planeplot = scatterplot3d(dt$x,dt$y,dt$z, angle=35,
  xlab = "x", ylab = "y", zlab = "z")
planeplot$plane3d(plane)
par(mfrow=(c(1,1)))
# dev.off()
```


```{r terraingGGplot, include=F, eval=F}
# ggplot(griddf, aes(x, y, z = z)) +
#   geom_contour_filled(binwidth = 30) +
#   geom_contour(bins = 50) +
#   geom_point(data = dt, shape=3) +
#   coord_fixed() +
#   #scale_fill_brewer(palette = "terrain") +
#   labs(title = "Homemade with GG") +
#   theme_minimal()
```



\textcolor{green}{Explain what we see and answer if a stationary grf is a suitable model}

Here we see a decreasing trend in positive $y$ direction, which suggest a non-constant mean function $m$. Thus, a stationary GRF model may not be suitable.

## b) Universal Kriging
Let $X = (X(s_1),...,X(s_{52}))^T$ be the vector of exact observations at locations $s_i,\quad i\in[1,52]$. Further assume that $X$ is a GRF modeled by
$$
\begin{split}
  E[X(s)] &= g(s)^T\beta \\
  Var[X(s)] &= \sigma^2 \\
  Corr[X(s), X(s')] &= \rho(||s - s'||), s,s' \in \mathcal{D}.
\end{split}
$$
Since correlation function is clearly isotropic, then so is the covariance function. \textcolor{gray}{(means invariant to shift and rotation. Furthermore, cov fnc is the stationary cov fnc. This is a powered exponential model.)} 

\textcolor{blue}{refer to text setup or write in full as done below?}

By following the notation in Spatial Statistics and Modeling[^1], where

[^1]:Gaetan & Guyon (2009), p.44. not sure if needed, but an example of footnote.

$$
\begin{split}
  \vec{z_i} &= (g_1(\vec{s_i}), ..., g_p(\vec{s_i}))^T \\
  Z &= [z_1^T, ..., z_{52}^T]^T \\
  z_0 &= (g_1(\vec{s_0}),..., g_p(\vec{s_0}))^T,
\end{split}
$$


1. 
$$
\hat{\vec{\beta}} = (Z^T \Sigma^{-1}Z)^{-1} Z^T \Sigma \vec X
$$
2. 
$$
\hat{\vec{X_0}} = \vec z_0 \hat{\vec{\beta}} + \vec c^T \Sigma^{-1} (\vec X-Z\hat{\vec{\beta}})
$$

\textcolor{blue}{c is covar fnc mentioned in task 1. Need to mention here?}
$$
\begin{split}
  \Sigma &= Cov(\vec X) = \left( 
  \begin{array}{rl}
    \Sigma_{00} & \Sigma_{01} \\
    \Sigma_{10} & \Sigma_{11}
  \end{array} \right) \\
  \vec c &= Cov(\vec X, X_0) = \sigma^2\rho?? \\
  \rho(h) &= \exp(-(0.05)^{1.5}) 
\end{split}
$$

\textcolor{green}{universal Kriging prediction variance. Does $\sigma^2$ need to change for different parameterizations of the expectation function?} 

\textcolor{red}{Written stuff in overleaf}

## c) Ordinary kriging
Model as in universal kriging, but with mean $E[X(s)] = g(s)\beta_1 = \beta_1 \Leftrightarrow g(s) = 1,\quad \forall s\in \mathcal{D}$. For the code we chose the range by where the correlation function is $\rho(h)=0.1 \Leftrightarrow h = 174$ (typically 10-13%)
```{r 2c_ordKrig, options}
D = seq(1,315,1)
gridOrd = expand.grid(x = D, y = D)
# Spatial Prediction - Ordinary Kriging
spOK = krige.conv(
  geodata=gd,
  locations=gridOrd,
  krige = krige.control(
    # defaults to ordinary Kriging and constant mean
    cov.model = "powered.exponential",
    cov.pars = c(50^2, 100), # (partial sill, range)
    kappa = 1.5 # Smoothness
    )
)
```

```{r 2c_ordKrig2, options}
# Beta estimate
spOK$beta.est
# manual


predOK = matrix(spOK$predict, ncol = 315) # to fit image.plot structure
varOK = matrix(spOK$krige.var, ncol = 315)

# Kriging predictor
# pdf("2c_ordKrigPred.pdf")
predImageOK = list(x = D, y = D, z = predOK)
image.plot(predImageOK, col=terrain.colors(200),asp=1)
contour(predImageOK, add = T)
# dev.off()

# Associated prediction variance
# pdf("2c_ordKrigVar.pdf")
varImageOK = list(x = D, y = D, z = varOK)
image.plot(varImageOK, col=viridis(200),asp=1)
points(dt$x, dt$y, col="white", pch=4)
# dev.off()


# surface3d = s3dPlot(predImageOK, fn = "2c_ordKrigSurface.pdf")
```


```{r diffTerrain, include = F, eval=F}
# grid2 = interp(x,y,z, nx=315,ny=315)
# D = seq(1,315,1)
# gridOrd = expand.grid(x = grid2$x, y = grid2$y)
# gd = as.geodata(dt)
# typeof(gd$coords)
# phi = 174 # Range set st. rho(phi) = 0.1
# # Spatial Prediction - Ordinary Kriging
# spOK = krige.conv(coords=cbind(x=x,y=y), data=z, locations=gridOrd,
#            krige = krige.control(
#              # defaults to ordinary Kriging and constant mean
#              cov.model = "powered.exponential",
#              cov.pars = c(50^2, 174), #(partial sill, range)
#              kappa = 1.5 # Smoothness
#            )
#            )
# par(mfrow=c(1,2))
# test = abs(grid2$z - matrix(spOK$predict, ncol = 315))
# image.diff = list(x = grid2$x, y=grid2$y, z=test)
# image.plot(image.diff, col=terrain.colors(52), asp=1, main = "Diff")
# points(dt, pch=4, col="red")
# 
# xyp <- as.ppp(cbind(x,y), c(0, 315, 0, 315))
# plot(density(xyp), main = "observation intensity")
# par(mfrow=c(1,1))
```

\textcolor{green}{Comment on results.}

## d) Universal Kriging
We denote the site $s = (s_1, s_2)$, and we set $n_g = 6$ to be the total number of known functions in the vector-valued function $g$. Furthermore, we specify each function in $g$ to be all polynomials $s_1^k s_2^l$ for $(k,l)\in{(0,0), (1,0), (0,1),(1,1), (2,0), (0,2)}$. Thus, the $6$-dimensional vector is \textcolor{gray}{use array instead}

\begin{equation}
  g(s) = [1, s_1, s_2, s_1s_2, s_1^2, s_2^2]
\end{equation}

The expected value of $X(s)$ is then

\begin{equation}
E[X(s)] = g(s)^T\beta = \beta_1 + \beta_2 s_1 + \beta_3 s_2 \beta_4s_1s_2 + \beta_5 s_{1}^{2} + \beta_6 s_{2}^{2}
\end{equation}


```{r 2d_uniKrig, options}
gridUniv = expand.grid(x = D, y = D)
spUK     = krige.conv(
  geodata=gd,
  # coords    = cbind(x=x,y=y), 
  # data      = z, 
  locations = gridOrd,
  krige     = krige.control(
    type.krige = "ok",
#"2nd": beta0 + beta1*x1 + beta2*x2 + beta3*(x1)^2 + beta4*(x2)^2 + beta5*x1*x2
    trend.d    = "2nd", 
    trend.l    = "2nd",
    cov.model  = "powered.exponential",
    cov.pars   = c(50^2, 100),
    kappa      = 1.5
  )
)
```
```{r 2d_uniKrigPreds}
# Estimated beta Universal Kriging
betaHat = (spUK$beta.est)
o2 = c(1,2,3,6,4,5) # order of betas corresponding to task description
betaHat[o2]

# Universal Kriging predictor
predUK = matrix(spUK$predict, ncol=315)
imageUK = list(x=D, y=D, z=predUK)
# pdf("2d_univKrigPred.pdf")
image.plot(imageUK,
           col=terrain.colors(200), 
           asp=1)
contour(list(x=D, y=D, z=predUK), add=T)
# points(dt$x,dt$y, pch=4, col = "red")
# dev.off()

# Associated variance predictions
varUK = matrix(spUK$krige.var,ncol = 315) # Predicted variance
# pdf("2d_univKrigVar.pdf")
image.plot(list(x=D,y=D,z=varUK), col=viridis(200), asp=1)
points(dt$x,dt$y, pch=4, col = "white")
# dev.off()
# surface3d2 = s3dPlot(imageUK, fn="2d_univKrigSurface.pdf")

# Difference ordinary and universal
predDiff = (predUK - predOK)
pdf("2d_predDiff.pdf")
image.plot(list(x=D,y=D,z=predDiff), col=turbo(200), asp=1)
# points(dt$x,dt$y, pch=4, col = "white")
dev.off()
# Logarithmic color gradient
predDiffLog = abs(predUK - predOK)
breaksPred = 1.1^(-190:10)/1.1^10*max(predDiff)
pdf("2d_predDiffLog.pdf")
image.plot(list(x=D,y=D,z=predDiffLog), col=viridis(200), asp=1,breaks = breaksPred)
# points(dt$x,dt$y, pch=4, col = "white")
dev.off()

varDiff = varUK - varOK
breaksVar = 1.1^(-189:10)/1.1^10*(max(varDiff)+1)
breaksVar = c(-0.001,breaksVar)
pdf("2d_varDiff.pdf")
image.plot(list(x=D,y=D,z=varDiff), col=turbo(200), asp=1, breaks = breaksVar)
# points(dt$x,dt$y, pch=4, col = "white")
dev.off()
image.plot(list(x=D,y=D,z=varDiff), col=turbo(200), asp=1)
```

## e) 
```{r 2e_2cReKrige, options}
D = seq(1,315,1)
gridUniv = expand.grid(x = D, y = D)
spUK     = krige.conv(geodata = gd,
  locations = gridUniv,
  krige     = krige.control(
    type.krige = "ok",
    #"2nd": beta0 + beta1*x1 + beta2*x2 + beta3*(x1)^2 + beta4*(x2)^2 + beta5*x1*x2
    trend.d    = "2nd",
    trend.l    = "2nd",
    cov.model  = "powered.exponential",
    cov.pars   = c(50^2, 100),
    kappa      = 1.5
  )
)

```

```{r 2e_plots, options}

predUK = matrix(spUK$predict,ncol = 315)

```


```{r 2e_probs, options}
# deleteResults = c(q1=
#   pnorm(850, mean = r0_hat, sd = sqrt(sigma_r_hat2), lower.tail = FALSE),
#   q2=
#   qnorm(0.9, mean = r0_hat, sd = sqrt(sigma_r_hat2))
# )


vars0 = matrix(spOK$krige.var, ncol=315)[100,100]
x0 = matrix(spOK$predict, ncol = 315)[100,100]
s0 = predOK[100,100]
c(krigPred = x0, predVar = vars0)
# P(X((100,100))>850)
pnorm(850, mean = x0, sd=sqrt(vars0), lower.tail = F)
# critical value for which the true val is below
# with a prob of 0.9
qnorm(0.9, x0, sqrt(vars0))
```

We have assumed normally distributed errors $\varepsilon$ for the linear regression model. Thus, our predictions $\hat X(s_o) = \hat X((100,100)^T)$ are normally distributed about the mean $E\hat X_0$. Considering the unobserved location $s_0 = (100,100)^T$ we get the Kriging predictor $\hat X((100,100)^T) \approx 851$ with associated variance $\sigma^2_{\bs{\hat X}}$


 